On the Minimum Core Mass for Giant Planet Formation

### 2013-10-20 tex/draft/draft4.tex
updated abstract, summary, and minor edits throughout

The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs. This core mass is typically quoted to be around $10 M_{\oplus}$. In static calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. In this study we consider the limiting case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet. We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk lifetime for a variety of disk conditions. We denote this mass as  \textit{critical core mass}. The absolute minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for an ideal diatomic gas with a solar composition and a standard ISM dust opacity. Lower opacity and disk temperature significantly reduce the critical core mass, while a decrease in the mean molecular weight of the nebular gas results in a larger critical core mass. Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. As additional heating due to planetesimal accretion leads to slower cooling, our results represent an absolute minimum for the core mass needed to form a giant planet before disk dissipation.

@@

To model the growth of planetary atmospheres around a solid core, we develop a simplified two layer model for time-dependent atmospheric cooling, i.e. KH contraction.  With a convective interior and radiative exterior, this model is motivated by similar models of hot Jupiters \citep{ab06, ym10}.

@@

We adopt a minimum mass solar nebula (MMSN) model for a passively irradiated disk \citep{chiang10}. With the semi-major axis $a$ normalized to the outer disk as $\aun{10} = a/(10 \text{ AU})$, the gas surface density and mid-plane temperature are  
\begin{subeqnarray} \label{eq:diskparam}
\varSigma\di  &=& 70 \,F_\varSigma \aun{10}^{-3/2} ~{\rm g~cm}^{-2} \\
T\di &=& 45  \,F_T\, \aun{10}^{-3/7} ~{\rm K} \, .
%\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
%T\di &=& 120 F_T a^{-3/7} \, \text{K}, \slabel{eq:diskb}
\end{subeqnarray}
The normalization factors $F_{\Sigma}$ and $F_T$  adjust the model relative to the fiducial MMSN.  We fix $F_{\varSigma}=F_T=1$ unless noted otherwise.

@@

To simplify our calculations of atmospheric contraction we use a two layer model with a bottom convective region and an upper radiative layer.   The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools, the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However, the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:op}.

@@

To make a single static atmosphere model (indexed by $i$), we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  To integrate Equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.  The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.  The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.  Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.

@@

\subsection{Two Layer Structure}
To apply the two layer cooling model analytically, we require expressions for the atmospheric structure.  Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses. (Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as 
\begin{subeqnarray}\label{eq:cb2}
T\cb &=& \chi T\di \slabel{eq:Tcb} \\
P\cb &=& \theta P_{\rm d} e^{R_{\rm B}/R\cb} \, .\slabel{eq:PcbRcb}
\end{subeqnarray}
The leading constants would be unity, $\chi = \theta = 1$, if the radiative zone were replaced by an isothermal layer.  In practice, deviations from unity are modest.    Standard radiative structure calculations (see \App{RCBcorr} for details) give 
\begin{equation}
\label{eq:chi}
\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} \simeq 1.53 \, ,
\end{equation}
for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at infinite depth, $\nabla_\infty = 1/2$, for our dust opacity.  The radiative zones are thus nearly isothermal, as also found by R06.  While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.   
%According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.

@@

Deep in the adiabatic interior, where $r \ll R\cb \ll \RB'$, the profiles follow simple power laws,
\begin{subeqnarray}\label{eq:deep}
\rho &\simeq& \rho\cb \left[ {\RB' \over r}\right]^{1/(\gamma -1)} \propto r^{-5/2} \slabel{eq:rhoconvdeep}  \\
T	&\simeq& T\cb {\RB' \over r} = {G M\co \over C_P r} \slabel{eq:Tconvdeep}\, .
\end{subeqnarray} 
While the radial density profile depends on the adiabatic index, the $r^{-1}$ temperature scaling is universal.  For reference, if self-gravity is included, the temperature profile satisfies
\begin{equation} \label{eq:TPsg}
{dT \over dr} = - {G m(r) \over C_P r^2}\, .
\end{equation} 
Thus in self-gravitating models the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$.

@@

Aside from this insight, the analytic model is useful because, despite the amplitude error, it explains the basic scaling of numerical cooling models.  As is well known, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass. Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.  Furthermore, the decline in both quantities with semi-major axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).  The temperature dependence is a competition between two main effects.  The higher luminosity, $\propto T^{4-\beta}$, at higher temperatures gives faster cooling, which opposes the overall effect.  The larger Bondi radius, and the effectively stronger gravity of the core, is the dominant effect that gives faster cooling at lower temperatures.  For different $\beta$ values, $t_{\rm co} \propto F_T^{ \beta + 1/2}$, so the slope of the dust opacity is an important moderating effect on the cooling.

The cooling timescale and critical core mass depend only weakly on the disk pressure, via the  logarithmic factor $\xi$.  Section \ref{sec:TPeffects} shows that these analytic predictions for disk temperature and pressure dependence compare favorably with numerical results.  %This weak dependence is a consequence of cooling coming from the radiative-convective boundary.  Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity. % and is here given the appropriate value for the nominal parameters.

\emph{TODO: consider adding "revised" analytic expressions for $t_{\rm co}$ and $\MC$ (e.g. lower Matm criterion).  People like scalings, but the obviously wrong amplitudes will prevent people from using them unless we correct.}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
%\vspace{-0.5in}
\caption{Similar to \Fig{fig:Ltplot}, but showing the effect of reducing the dust abundance by factors of 10 and 100 from standard Solar abundances. The reduced dust opacities give higher luminosities and faster atmospheric growth.}  %The actual magnitude of dust opacities is difficult to robustly predict.}
\label{fig:LtvsMopacity}
\end{figure}

@@

At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as negative time steps when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.  Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after the phase of slowest growth.  The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.   Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.

@@

A second radiative zone inside the convection zone, a ``radiative window," is possible, due to the opacity drops caused by the evaporation of ice and metal grains. Our two layer model ignores such complications. Grain growth also impacts the opacity in the outer regions as well as the existence or depth of radiative windows. We explore more realistic opacity laws in  In Piso et al. (2013, in prep.).

@@

 In this paper we study the formation of giant planet atmospheres under the assumption of a low planetesimal accretion regime in which KH contraction dominates the luminosity evolution of the atmosphere. We consider atmosphere growth around a core of fixed mass, and we build quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk at the Hill radius. In our static models, we use an ideal gas polytropic EOS, and assume a standard ISM power-law opacity for dust grains. We derive a cooling model to connect series of quasti-static atmospheres and obtain an evolutionary history of the envelope. We define the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, for which $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$. From this we define as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk. We calculate the critical core mass for a variety of disk conditions, nebular gas compositions and opacities. The critical core mass is $\sim 8 M_{\oplus}$ at 5 AU in our fiducial disk and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU. A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low.

Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies. A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime. Moreover, we are able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool.

>>>>

The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. This value, however, depends on several factors, such as the location in the disk, nebular gas composition, or the accretion rate of solids. In this study we consider the limiting case of atmosphere growth around a core of fixed mass, in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the envelope. We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within typical disk life timescales for a variety of disk conditions. We are primarily interested in the outer regions of the protoplanetary disk, where our assumptions of spherical symmetry and dust grains opacity are valid, and where the increasing number of directly imaged giant planets helps constrain formation models. The absolute minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for an ideal diatomic gas with a solar composition and a standard ISM dust opacity. Lower opacity and disk temperature significantly reduce the critical core mass, while a decrease in the mean molecular weight of the nebular gas results in a larger critical core mass. Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. As additional heating due to planetesimal accretion leads to slower cooling, our results represent an absolute minimum for the core mass needed to form a giant planet before disk dissipation.

%Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs. This core mass is typically quoted to be around $10 M_{\oplus}$. In static calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. In this study we consider the limiting case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.

@@

To model the growth of planetary atmospheres around a solid core, we develop a simplified two-layer model for time-dependent atmospheric cooling, i.e. KH contraction.  With a convective interior and radiative exterior, this model is motivated by similar models of hot Jupiters \citep{ab06, ym10}.

@@

We adopt a minimum mass solar nebula (MMSN) model for a passively irradiated disk \citep{chiang10}. With the semimajor axis $a$ normalized to the outer disk as $\aun{10} = a/(10 \text{ AU})$, the gas surface density and mid-plane temperature are  
\begin{subeqnarray} \label{eq:diskparam}
\varSigma\di  &=& 70 \,F_\varSigma \aun{10}^{-3/2} ~{\rm g~cm}^{-2} \\
T\di &=& 45  \,F_T\, \aun{10}^{-3/7} ~{\rm K} \, .
%\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
%T\di &=& 120 F_T a^{-3/7} \, \text{K}, \slabel{eq:diskb}
\end{subeqnarray}
The normalization factors $F_{\Sigma}$ and $F_T$  adjust the model relative to the fiducial MMSN.  We fix $F_{\varSigma}=F_T=1$ unless noted otherwise.

@@

To simplify our calculations of atmospheric contraction we use a two-layer model with a bottom convective region and an upper radiative layer.   The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools, the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However, the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:op}.

@@

To obtain a single atmosphere profile (indexed by $i$), we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  To integrate Equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.  The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.  The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.  Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.

@@

\subsection{Two Layer Structure}
To apply the two-layer cooling model analytically, we require expressions for the atmospheric structure.  Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses. (Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as 
\begin{subeqnarray}\label{eq:cb2}
T\cb &=& \chi T\di \slabel{eq:Tcb} \\
P\cb &=& \theta P_{\rm d} e^{R_{\rm B}/R\cb} \, .\slabel{eq:PcbRcb}
\end{subeqnarray}
The leading constants would be unity, $\chi = \theta = 1$, if the radiative zone were replaced by an isothermal layer.  In practice, deviations from unity are modest.    Standard radiative structure calculations (see \App{RCBcorr} for details) give 
\begin{equation}
\label{eq:chi}
\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} \simeq 1.53 \, ,
\end{equation}
for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at infinite depth, $\nabla_\infty = 1/2$, for our dust opacity.  The radiative zones are thus nearly isothermal, as also found by R06.  While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.   
%According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.

@@

Deep in the adiabatic interior, where $r \ll R\cb \ll \RB'$, the profiles follow simple power laws,
\begin{subeqnarray}\label{eq:deep}
\rho &\simeq& \rho\cb \left[ {\RB' \over r}\right]^{1/(\gamma -1)} \propto r^{-5/2} \slabel{eq:rhoconvdeep}  \\
T	&\simeq& T\cb {\RB' \over r} = {G M\co \over C_P r} \slabel{eq:Tconvdeep}\, .
\end{subeqnarray} 
While the radial density profile depends on the adiabatic index, the $r^{-1}$ temperature scaling is universal.  For reference, if self-gravity is included, the temperature profile satisfies
\begin{equation} \label{eq:TPsg}
{dT \over dr} = - {G m(r) \over C_P r^2}\, .
\end{equation} 
Thus, the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$ in self-gravitating models.

@@

Aside from this insight, the analytic model is useful because, despite the amplitude error, it explains the basic scaling of numerical cooling models.  As is well known, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass. Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.  Furthermore, the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).  The temperature dependence is a competition between two main effects.  The higher luminosity, $\propto T^{4-\beta}$, at higher temperatures gives faster cooling, which opposes the overall effect.  The larger Bondi radius, and the effectively stronger gravity of the core, is the dominant effect that gives faster cooling at lower temperatures.  For different $\beta$ values, $t_{\rm co} \propto F_T^{ \beta + 1/2}$, so the slope of the dust opacity is an important moderating effect on the cooling.

The cooling timescale and critical core mass depend only weakly on the disk pressure, via the  logarithmic factor $\xi$.  Section \S\ref{sec:TPeffects} shows that these analytic predictions for disk temperature and pressure dependence compare favorably with numerical results. Moreover, the analytic estimates quantitatively reproduce the numerical results to reasonable accuracy, if the effective crossover mass for unstable atmosphere collapse is a fraction of the core mass, i.e. $M_{\rm atm}=f M\co$. Under this assumption, Equations (\ref{eq:tcoolan}) and (\ref{eq:Mcrit}) become
\begin{eqnarray} 
\label{eq:tcoolanscaled}
t_{\rm co} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2 f^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr \\ \slabel{eq:tcoolanf}
 & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr
\end{eqnarray} 
and
\begin{eqnarray}\label{eq:Mcritscaled}
\MC &\approx & 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} f^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus \\ \slabel{eq:Mcritf}
& \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ ,
\end{eqnarray}
with $f=0.13$ in Equations (\ref{eq:tcoolanf}) and (\ref{eq:Mcritf}). We choose this particular fractional scaling to match our numerical results at 10 AU (see section \S\ref{sec:critical} for more details). We note that the scaling above is valid if the logarithmic factor $\xi$ is fixed; however, the analytic estimates presented in section \S\ref{sec:critical} explicitly take into account the dependence of $\xi$ on the core mass (see Equations \ref{eq:xidef} and \ref{eq:Pcbc}).

 %This weak dependence is a consequence of cooling coming from the radiative-convective boundary.  Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity. % and is here given the appropriate value for the nominal parameters.

%\emph{TODO: consider adding "revised" analytic expressions for $t_{\rm co}$ and $\MC$ (e.g. lower Matm criterion).  People like scalings, but the obviously wrong amplitudes will prevent people from using them unless we correct.}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
%\vspace{-0.5in}
\caption{Similar to \Fig{fig:Ltplot}, but for a $5 M_{\oplus}$ core forming at 10 AU; we show the effect of reducing the dust abundance by factors of 10 and 100 from standard Solar abundances. The reduced dust opacities give higher luminosities and faster atmospheric growth.}  %The actual magnitude of dust opacities is difficult to robustly predict.}
\label{fig:LtvsMopacity}
\end{figure}

@@

At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as negative time steps when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.  Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after growth has started to accelerate. In our models the breakdown occurs at lower atmosphere mass as the semimajor axis increases, since accelerated and eventually runaway growth occurs sooner at larger distances, as seen in \Fig{fig:growthtime}. The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.   Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.

@@

A second radiative zone inside the convection zone, a ``radiative window," is possible, due to the opacity drops caused by the evaporation of ice and metal grains. Our two-layer model ignores such complications. Grain growth also impacts the opacity in the outer regions as well as the existence or depth of radiative windows. We explore more realistic opacity laws in  In Piso et al. (2013, in prep.).

@@

%In this paper we study the formation of giant planets through core accretion.

In this paper we build evolutionary models of giant planets forming in a protoplanetary disk. We consider atmosphere growth around a core of fixed mass, and determine the minimum core mass for which unstable atmosphere collapse occurs before the dissipation of the protoplanetary disk, at various locations in the disk. For a standard ISM power-law opacity and a polytropic EOS for the nebular gas, the minimum core mass is $\sim 8 M_{\oplus}$ at 5 AU and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU. A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low.  Our results are applicable in the outer regions of the disk ($\gtrsim 5$ AU), where planets can undergo runaway gas accretion before opening a gap in the disk. If follows that core accretion at wide separations may result in massive gas giants, which could explain the formation of some of the directly imaged giant planets (\citealt{marois08}; \citealt{kalas08}; \citealt{lagrange10}).

Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies. A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime. We note that our model is a limiting case of core accretion, in which planetesimal accretion is negligible and the evolution of the gaseous envelope is dominated by KH contraction. We are thus able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool. Our results are not inconsistent with the existence of ice giants at large distances (such as Uranus and Neptune in our own Solar System), as ongoing accretion of solids can stall atmospheric growth and prevent the accumulation of a massive envelope before disk dissipation.

% In this paper we study the formation of giant planet atmospheres under the assumption of a low planetesimal accretion regime in which KH contraction dominates the luminosity evolution of the atmosphere. We consider atmosphere growth around a core of fixed mass, and we build quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk at the Hill radius. In our static models, we use an ideal gas polytropic EOS, and assume a standard ISM power-law opacity for dust grains. We derive a cooling model to connect series of quasti-static atmospheres and obtain an evolutionary history of the envelope. We define the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, for which $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$. From this we define as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk. We calculate the critical core mass for a variety of disk conditions, nebular gas compositions and opacities. The critical core mass is $\sim 8 M_{\oplus}$ at 5 AU in our fiducial disk and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU. A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low. 
% 
%Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies. A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime. Moreover, we are able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool.
